---
output:
  github_document
---
[![Build Status](https://travis-ci.org/lauzikaite/massFlowR.svg?branch=master)](https://travis-ci.org/lauzikaite/massFlowR)

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      eval = FALSE, 
                      message =  TRUE)
massFlowR_url <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/massFlowR.html"
processing_url <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/processing.html"
annotation_url <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/annotation.html"

```

# massflowR

Package for pre-processing of high-resolution, untargeted, centroid LC-MS data.


`massFlowR` detects and aligns structurally-related spectral peaks across LC-MS experiment samples.


***

## Individual samples processing


Each LC-MS file in the experiment is processed independently.
 
* Chromatographic peak detection enabled by the *centWave* algorithm from `xcms` package.

* Grouping structurally-related chromatographic peaks into **pseudo chemical spectra** (PCS).

Individual samples can be processed in real-time during LC-MS spectra acquisition, or in parallel, using backend provided by `doParallel` package.


## Peak alignment


To align chromatographic peaks across all samples in LC-MS experiment, an algorithm, which compares the overall similarity of pseudo chemical spectra is implemented (see [Peak alignment](`r processing_url`)). 


## Post-alignment processing


Once pseudo chemical spectra are aligned across samples, the obtained features are validated. Intensity values for each feature in a PCS are correlated across all samples to identify sets of features with similar correlation pattern.

Final step in the pipeline is to re-integrate intensity values for chromatographic peaks that were not detected by the *centWave* using raw LC-MS files. *m/z* and *rt* integration regions are estimated for each sample separetely through cubic smoothing spline interpolation.

In the generated feature table, each feature is assigned to a pseudo chemical spectra, which can be used in manual metabolite annotations.


## Automatic annotation


If in-house chemical reference database is available, obtained PCS are annotated. For more details how to build a database file, see [annotation using database](`r annotation_url`)).


***


## Vignettes

More information is available in vignettes:

* [massFlowR overview](`r massFlowR_url`)
* [Data processing](`r processing_url`)
* [Automatic annotation](`r annotation_url`)

***


# Installation


```{r}
# To install dependencies
install_dependencies <- function () {
  dependencies <- c("xcms",  "MSnbase", "faahKO", "igraph", "doParallel",  "foreach")
  installed <- installed.packages()
  to_install <- subset(dependencies, !(dependencies %in% installed[, "Package"]))
  if (length(to_install) != 0) {
    if (!requireNamespace("BiocManager", quietly = TRUE)) {
      install.packages("BiocManager")
    }
    message("Installing packages: ", to_install, " ...")
    BiocManager::install(to_install, version = "3.8")
  } else {
    message("All dependencies are already installed.")
  }
}
install_dependencies()

# To install development version
devtools::install_github("lauzikaite/massflowR", dependencies = FALSE)
```

