---
output:
  github_document
---

```{r, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      collapse = TRUE,
                      eval = FALSE, 
                      message =  TRUE)
url <- "http://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/doc/massFlowR.html"
```

# massflowR

Package for pre-processing of high-resolution, untargeted, centroid LC-MS data.


`massFlowR` detects and aligns structurally-related spectral peaks across LC-MS experiment samples.


***

## Individual samples processing


Each LC-MS file in the experiment is processed independently.
 
* Chromatographic peak detection enabled by the *centWave* algorithm from `xcms` package.

* Grouping of chromatographic peaks into structurally-related groups of peaks (see [Peak Grouping](`r url`)).

Individual samples can be processed in real-time during LC-MS spectra acquisition, or in parallel, using backend provided by `doParallel` package.


## Peak alignment


To align peaks across all samples in LC-MS experiment, an algorithm, which compares the overall similarity of structurally-related groups of peaks is implemented (see [Peak alignment](`r url`)). 


## Post-alignment processing


Once samples are aligned, the obtained peak groups are validated. Intensity values for each peak in a group are correlated across all samples to identify sets of peaks exhibiting similar behaviour. These sets of peaks are called **Pseudo Chemical Spectra** (PCS).

Final step in the pipeline is to re-integrate intensity for peaks that were not detected by the *centWave* using raw LC-MS files. *m/z* and *rt* values for intensity integration are estimated for each sample separetely. *m/z* and *rt* values are intrapolated using a cubic smoothing spline.


## Peak annotation


If in-house chemical reference database is available, PCS are annotated. For more details how to build a database file, see [annotation using database](`r url`)).


***


# Installation


```{r}
# For devel version
devtools::install_github("lauzikaite/massflowR")
```

