---
title: "Annotation using chemical reference database"
output:
  BiocStyle::html_document:
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Annotation using chemical reference database}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{massFlowR}
  %\VignetteKeywords{mass spectrometry, metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

**Package**: [massFlowR](https://github.com/lauzikaite/massFlowR)<br />
**Authors**: Elzbieta Lauzikaite<br />
**Modified**: `r file.info("massFloWR.Rmd")$mtime`<br />
**Compiled**: `r date()`

```{r  setup, echo = FALSE, results = "asis" }
BiocStyle::markdown() 
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
Biocpkg <- function (pkg) {
    sprintf("[%s](http://bioconductor.org/packages/%s)", pkg, pkg)
}
options(kpb.suppress_noninteractive = TRUE) ## supressing progress bar
```

```{r libraries, message = FALSE, echo = FALSE}
suppressWarnings(library(massFlowR))
suppressWarnings(library(xcms))
suppressWarnings(library(dplyr))
rda_dir <- "~/Documents/Projects/DBannotate/NPC-RPOS-LCE-SubDB"
db_dir <- "~/Documents/Projects/massFlowR/packageDEV/db"
out_directory <- getwd()
url <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/massFlowR.html"
```


`massFlowR` performs automatic annotation during peak alignment if LC-MS files for chemical reference compounds are available. To enable annotation, provide a database table in a csv format when initiating peak alignment (see section [peak annotation](`r url`)).

Database table can be obtained from raw LC-MS files in two steps:

* Build pseudo chemical spectra (PCS) for each compound using the raw LC-MS file (in mzML/NetCDF format)
* Build database table


# Chemical spectra extraction


*(code under development)*

Raw LC-MS files had been processed by M.Lewis. Code for corresponding functionality will be added to `massFlowR` package for those that have acquired LC-MS data independently. 


# Database table


## Building database from rda files


Previously processed LC-MS files were recorded in rda files. Function `getDB` can be used to build a database table using these rda files. The generated table will have the following columns:

* `peakid` (unique peak number)
* `mz` (peak m/z)
* `rt` (peak retention time, sec)
* `into` (peak intensity)
* `peakgr` (unique peak-group, PCS, number)
* `chemid` (unique database chemical number)
* `dbid`  (compound identifier)
* `dbname` (compound chemical name)


```{r db_table, message = FALSE, echo = FALSE, results = 'hide'}
buildDB(rda_dir = rda_dir, out_dir = out_directory)
db_table <- read.csv(file.path(out_directory, "DBtemplate.csv"))
```

```{r db_table_visible, eval=FALSE}
# rda_dir <- "absolute_path_to_rda_files"
# out_dir <- "absolute_path_to_output_directory"
buildDB(rda_dir = rda_dir, out_dir = out_directory)
db_table <- read.csv(file.path(out_directory, "DBtemplate.csv"))
```

```{r db_table_knitr, echo = FALSE}
db_table <- db_table[which(db_table$chemid %in% c(1,2,3)),]
dt <- head(db_table)
knitr::kable(x = dt,
             format = "html",
             row.names = F) %>% 
  kableExtra::kable_styling(full_width = TRUE, bootstrap_options = c("striped"), position = "left")
```


# See also


* [Introduction to massFlowR](`r url`)
