---
title: "LC-MS data processing with massFlowR"
output:
  BiocStyle::html_document:
    toc_float: true
    collapsed: false
vignette: >
  %\VignetteIndexEntry{User-guide to massFlowR package}
  %\VignetteEncoding{UTF-8}
  %\VignettePackage{massFlowR}
  %\VignetteKeywords{mass spectrometry, metabolomics}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
bibliography: references.bib
references:
- id: dummy
  title: no title
  author:
  - family: noname
    given: noname
---

<style>
pre code, pre, code {
  white-space: pre !important;
  overflow-x: scroll !important;
  word-break: keep-all !important;
  word-wrap: initial !important;
}
</style>

**Package**: [massFlowR](https://github.com/lauzikaite/massFlowR)<br />
**Authors**: Elzbieta Lauzikaite<br />
**Date**: `r date()`

```{r  setup, echo = FALSE, results = "asis" }
BiocStyle::markdown() 
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
Biocpkg <- function (pkg) {
    sprintf("[%s](http://bioconductor.org/packages/%s)", pkg, pkg)
}
## supressing progress bar
options(kpb.suppress_noninteractive = TRUE) 
```

```{r libraries, message = FALSE, echo = FALSE}
## load libraries quietly to avoid printing messages in the vignette
suppressWarnings(library(massFlowR))
out_directory <- getwd()
url_m <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/massFlowR.html"
url_a <- "https://htmlpreview.github.io/?https://github.com/lauzikaite/massFlowR/blob/master/doc/annotation.html"
```


# Introduction


This documents provides an overview of the LC-MS data pre-processing with `massFlowR` using dataset faahKO.  


# Data import


LC-MS data in mzML/NetCDF format import is supported. Data import is implemented via `r Biocpkg("mzR")` package.

In this document, functionality of the package will be demonstrated using data from `r Biocpkg("faahKO")` package. Raw LC-MS data files (in NetCDF format) are provided for spinal cords samples taken from six knock-out (KO) and six wild-type (WT) mice. Each datafile contains centroided data acquired in positive ionization mode, with data recorded at 200-600 m/z and 2500-4500 seconds.

Load the package and locate the raw CDF files within the `faahKO` package:

```{r data_import}
library(massFlowR)
## Get the full path to the CDF files
faahKO_files <- dir(system.file("cdf/WT", package = "faahKO"), full.names = TRUE)
```


```{r data_import_knitr, echo = FALSE}
kableExtra::kable_styling(
  knitr::kable(x = faahKO_files,
             format = "html",
             row.names = FALSE, col.names = ""),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


# Individual samples processing

The first stage in the pipeline is chromatographic peak detection and grouping via function `groupPEAKS`.


## Chromatographic peak detection

Peaks are detected using the *centWave* algorithm from `r Biocpkg("xcms")` package (see [@Tautenhahn2008]). Appropriate parameters for the LC-MS experiment must be selected. For advise on this, please see the official `xcms` [manual](https://bioconductor.org/packages/release/bioc/vignettes/xcms/doc/xcms.html#3_initial_data_inspection). Selected parameters must be built into a *CentWaveParam* class object:


```{r params}
## xcms parameters for peak-picking
cwt_param <- xcms::CentWaveParam(ppm = 25,
                                 snthresh = 10,
                                 noise = 1000,
                                 prefilter =  c(3, 100),
                                 peakwidth = c(30, 80),
                                 mzdiff = 0)
```


## Chromatographic peak grouping


Detected peaks are put into groups, which comprise peaks originating from the same chemical compound: adducts and isotopes. For each peak in a sample, function `groupPEAKS`:

* Finds co-eluting chromatographic peaks;
* Performs extracted ion chromatogram (EIC) correlation between all co-eluting peaks;
* Builds a network of peaks with high EIC correlation;
* Detects communities of peak within the correlation network (implemented by `igraph` package algorithm, see [@Raghavan2007]).

Peaks that group into a community form a **pseudo chemical spectra**. Only communities with more than one peak are retained for further processing.


## Implementation


Function `groupPEAKS` processes every LC-MS datafile independently and thus can be implemented in parallel, or during sample acquisition on the machine linked to the LC-MS. A list of paths to LC-MS datafiles (in mzML/NetCDF format) is feeded to `groupPEAKS`, together with the *CentWaveParam* class object, path to output directory and parameters for parallelisation. 


`groupPEAKS` writes a csv with detected and grouped peaks in the selected directory for each LC-MS sample separately. The filenames of the generated csv files will be needed for the next stage in the pipeline. The filename starts with the original raw LC-MS filename and ends with "peakgrs.csv".


`massFlowR` pipeline requires a metadata table with the following columns for each sample:

* `filename` specifies the basename of the raw LC-MS file.
* `run_order` specifies the acquisition order for the corresponding LC-MS sample.
* `raw_filepath` specifies the absolute path to the raw LC-MS file (netCDF/mzML).


```{r metadata}
## define where processed datafiles should be written
# out_directory <- "absolute_path_to_output_directory"

## create metadata table with required columns 'filename', 'run_order' and 'raw_filepath'
metadata <-
  data.frame(
  filename = gsub(".CDF", "", basename(faahKO_files)),
  run_order = seq(length(faahKO_files)),
  raw_filepath = faahKO_files,
  stringsAsFactors = FALSE
  )
write.csv(metadata, file = file.path(out_directory, "metadata.csv"), row.names = FALSE)
```


```{r experiment_table, echo = FALSE}
kableExtra::kable_styling(
  knitr::kable(x = metadata,
             format = "html",
             row.names = FALSE),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


```{r peak_grouping}
## run peak detection and grouping for the listed faahKO files with two workers
groupPEAKS(file = file.path(out_directory, "metadata.csv"), out_dir = out_directory, cwt = cwt_param, ncores = 2)
```


# Peak alignment


To align structurally-related peaks as a group across samples in LC-MS experiment, an alignment algorithm, which preserves the structural spectral information, is implemented. 

Peaks are aligned by taking samples in the order of raw sample acquisition and matching them against a template. Template is list of all previously aligned peaks, which is updated with each sample by:

* adding new peaks
* averaging the *m/z* and *rt* values of matching peaks between the sample and the template.

Therefore, template stores the **moving averages** of *m/z* and *rt* values. 

For each peak in a sample, alignment algorithm:

1. Finds all template peaks within a *m/z* and *rt* window.
2. Identifies the true match by comparing the spectral similarity between the peak-group of the peak-of-interest and all matching template's peak-groups.
3. Merges the selected template's peak-group with the peak-group of the peak-of-interest. It updates template's *m/z* and *rt*  values for the matching peaks across the template and the sample.

Spectral similarity is measured by obtaining the cosine of the angle between two 2D vectors, representing each PCSs *m/z* and *intensity* values.


## Implementation


To enable peak alignment, previous metadata table has to contain additional column:

* `proc_filepath` specifies the absolute path to the csv files generated by the `groupPEAKS` function.


```{r metadata_update}
##  update previous metadata table and add paths to generated csv files
processed_files <- list.files(out_directory, "peakgrs.csv", full.names = TRUE)
metadata$proc_filepath <- processed_files
write.csv(metadata, file.path(out_directory, "metadata_grouped.csv"), row.names = FALSE)
```


```{r metadata_update_table, echo = FALSE}
kableExtra::kable_styling(
  knitr::kable(x = metadata,
             format = "html",
             row.names = FALSE),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


To initiate peak alignment, use function `buildTMP`, which constructs a `massFlowTemplate`class object. `massFlowTemplate` object stores sample alignment and annotation data and is updated with every sample. Define the desired error window for *m/z* and *rt* (seconds) values, which will be used for the whole experiment. `mz_err = 0.01` and `rt_err = 2` are recommended for high-resolution UPLC-MS data. `mz_err = 0.01` and `rt_err = 10` are suitable for the `faahKO`package data.

Desired spectral similarity threshold can be selected with *cutoff* parameter (0-1, with 1 being perfectly identical). `cutoff = 0.5` is recommended for high-resolution UPLC-MS data


```{r buildTMP}
## initiate template
template <- buildTMP(file = file.path(out_directory, "metadata_grouped.csv"), out_dir = out_directory, mz_err = 0.01, rt_err = 10, cutoff = 0.5)
```


To review the samples that are in the experiment, use slot `@samples`.


```{r samples_slot, eval = FALSE}
## review samples in the experiment using slot "samples"
template@samples
```


```{r samples_slot_knitr, echo = FALSE}
## review alignment results for an individual sample, e.g. the second, using slot "data"
dt <- template@samples
kableExtra::kable_styling(
  knitr::kable(x = dt,
             format = "html",
             row.names = F),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


`massFlowTemplate` object stores the most up-to-date template in the `@tmp` slot. Function `buildTMP` creates a template using the first sample in the experiment.


```{r tmp_slot, eval = FALSE}
## review gnerated template using slot "tmp"
head(template@tmp, 20)
```


```{r tmp_slot_knitr, echo = FALSE}
## review alignment results for an individual sample, e.g. the second, using slot "data"
dt <- head(template@tmp, 20)
kableExtra::kable_styling(
  knitr::kable(x = dt,
             format = "html",
             row.names = F),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


To align peaks across all samples in the study, apply method `alignPEAKS`. `alignPEAKS` updates the `massFlowTemplate` object:

1. Selects next sample to be aligned and checks whether it was already peak-picked and grouped (waits until the corresponding csv file is written, [see](#realtime-implementation)).
2. Matches every peak in the sample against the template.
3. Selects best matches using spectral similarity comparison.
3. Updates template with sample's peaks: adds new and averages matching peaks.

Parameter `ncores` allows a quicker implementation using the parallel backend that is available on the user's machine (i.e. multicore on Unix/Mac and snow on Windows). Select the desired number of parallel workers.


```{r alignPEAKS}
## align peaks across all remaining samples
template <- alignPEAKS(template, out_dir = out_directory, ncores = 2)
```


Aligned sample's peak tables are stored within the `@data` slot, which lists tables for each sample separately.


```{r data_slot, eval = FALSE}
## review alignment results for an individual sample, e.g. the second, using slot "data"
head(template@data[[2]])
```


```{r data_slot_knitr, echo = FALSE}
## review alignment results for an individual sample, e.g. the second, using slot "data"
dt <- head(template@data[[2]])
kableExtra::kable_styling(
  knitr::kable(x = dt,
             format = "html",
             row.names = F),
  full_width = TRUE, bootstrap_options = c("striped", "hover", "responsive"), position = "left")
```


## Alignment re-initiation


If alignment was stopped before all samples in the experiment could be aligned (for example, because of a computer issue), alignmment can be re-initiated by loading the samples that had already been processed. Function `loadALIGNED` requires the absolute paths to the updated metadata file and the final template writen by `alignPEAKS`. Note that same parameters must be provided as for the `buildTMP` function. 


```{r loadALIGNED1}
## get the absolute paths to the updated metadata file and the final template writen by "alignPEAKS" 
m_file <- file.path(out_directory, "aligned.csv")
tmp_file <- file.path(out_directory, "template.csv")

## load already aligned samples into a massFlowTemplate object
template <- loadALIGNED(file = m_file, template = tmp_file, mz_err = 0.01, rt_err = 10, cutoff = 0.5)
```


```{r loadALIGNED2}
## if some of the samples in the experiment have not been aligned yet, re-start alignment
template@samples$aligned
```


```{r loadALIGNED3}
if (any(template@samples$aligned == FALSE)) {
  template <- alignPEAKS(template, out_dir = out_directory, ncores = 2)
}
```

## Realtime implementation

Sample alignment can be performed in real-time during data acquisition by setting the `buildTMP` function parameter to `realtime = TRUE`. The function will wait for the intermediate peak-groups csv files to be written instead of stopping when a file that does not exist is encountered. Make sure, however, that correct filepaths are provided in the metadata csv file's column `proc_filepath`. 


```{r realtime, eval = FALSE}
template <- buildTMP(file = file.path(out_directory, "metadata_grouped.csv"), out_dir = out_directory, mz_err = 0.01, rt_err = 10, cutoff = 0.5, realtime = TRUE)
template <- alignPEAKS(template, out_dir = out_directory, ncores = 2)
```


# Alignment validation


Once peaks are aligned across all samples, the obtained peak-groups are validated. Intensity values for each peak in a group are correlated across all samples. Correlation estimates are then used to build networks of peaks, that behave similarly across all samples. Peaks exhibiting a different pattern in their intensities are put into a new peak-group.

Each peak-group is a *pseudo chemical spectra* (PCS), which comprised peaks exhibiting consistent behaviour across samples. 

To enable alignment validation, a metadata table and final template file that both were written by the `alignPEAKS` function in the selected directory, must be used. A `massFlowTemplate`class object, returned by the `alignPEAKS` function will be needed. If alignment validation will be performed in a new R session, re-initiate the `massFlowTemplate`class objec using `loadALIGNED` function as before. 

```{r loadALIGNED}
## get the absolute paths to the updated metadata file and the final template writen by "alignPEAKS" 
m_file <- file.path(out_directory, "aligned.csv")
tmp_file <- file.path(out_directory, "template.csv")

## initiate validation by first loading aligned samples into a massFlowTemplate object
template <- loadALIGNED(file = m_file, template = tmp_file)
```


Peak-group validation is enabled by applying the method `validPEAKS` on the `massFlowTemplate`class object. Validation can be implemented in parallel using `ncores` parameter. 


`validPEAKS` will return a `massFlowTemplate`class object with validated *pseudo chemical spectra*, as well as write peak tables for the obtained PCS:

* intensity_data.csv (intensity values for every peak in PCS in every sample)
* peaks_data.csv
* sample_data.csv


```{r validPEAKS}
## Start validation using a massFlowTemplate object
template <- validPEAKS(template, out_dir = out_directory, ncores = 2, cor_thr = 0.5)
```


# Peak filling


Final step in the pipeline is to re-integrate intensity values for peaks that were not detected by the *centWave* using raw LC-MS files. In contrast to `xcms` package, *m/z* and *rt* values for intensity integration are estimated for each sample separetely. *m/z* and *rt* values are modelled using a cubic smoothing spline.


```{r fillPEAKS}
## Fill peaks using validated massFlowTemplate object
template <- fillPEAKS(template, out_dir = out_directory, ncores = 2)
```


# Final output


`fillPEAKS` writes file 'filled_intensity_data.csv', which contains features metadata (including pseudo chemical spectra number) and intensity values for each sample in the experiment. 

```{r load_filled}
final_dt <- read.csv(file.path(out_directory, "filled_intensity_data.csv"), header = TRUE, stringsAsFactors = FALSE)
```


```{r ion_map, echo = FALSE}
ggplot2::ggplot(final_dt,
             ggplot2::aes(x = rt, y = mz, color = as.factor(pcs))) +
      ggplot2::geom_point(size = 3, alpha = 0.8) +
  ggplot2::xlab("Retention time") +
  ggplot2::ylab("m/z") +
  ggplot2::scale_color_viridis_d(name = "Pseudo chemical spectra") +
  ggplot2::theme_bw(base_size = 14) +
  ggplot2::theme(
    legend.position = "bottom",
    panel.grid.major = ggplot2::element_blank(),
    panel.grid.minor = ggplot2::element_blank(),
    axis.line = ggplot2::element_line(size = 0.1)
  )
```


# Peak annotation


If in-house chemical reference database is available, PCS are annotated. For more details how to build a database file, see [annotation using database](`r url_a`)).


# See also


* [massFlowR overview](`r url_m`)
* [Automatic annotation](`r url_a`)


```{r teardown, message = FALSE, echo = FALSE}
unlink(list.files(path = out_directory, pattern = ".csv", full.names = T))
unlink(list.files(path = out_directory, pattern = ".RDS", full.names = T))
```

# References